---
title: "classtpx on Ancient vs Modern damage data"
author: "Kushal K Dey"
date: "September 15, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

In this script, we observe whether the supervised topic model approach on Moderns and Ancients pooled data, taking moderns as the base is effective in uncovering structure in moderns and ancients data.



```{r}
library(slam)
library(maptpx)
library(devtools)
install_github("kkdey/classtpx")
library(classtpx)
library(CountClust)
```

## Loading signature counts data

```{r}
signature_counts <- get(load("../summary_data/signature-counts-clubbed-Lindo2016.rda"))
signature_set <- colnames(signature_counts)

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

```

## classtpx with moderns fixed

### K=2 

Applying classtpx with moderns as the base cluster to form supervised topic models. Use K=2.

```{r}
class_labs <- rep(1,25)
known_samples <- 26:50

Topic_clus <- class_topics(
  signature_counts, 
  K=2, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.99,0.01),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

### K=3

Use K=3.

```{r}
Topic_clus <- class_topics(
  signature_counts, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.90,0.01, 0.01),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

## classtpx with a uniform cluster 

### K=3

We use $K=3$, with one cluster having equal probability in all damage signatures.

```{r}
Topic_clus <- class_topics(
  signature_counts, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/length(signature_set), length(signature_set)),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.99,0.01,0.01),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

### K=4

We perform $K=4$ with one cluster fixed to have equal probability on all damage patterns.

```{r}
Topic_clus <- class_topics(
  signature_counts, 
  K=4, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/length(signature_set), length(signature_set)),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.99,0.01,0.01,0.01),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

## Removing C to T
We remove the  C to T substitutions, and then perform classtpx with moderns as the base population.

```{r}
indices <- which(sig_split[,3]=="C" & sig_split[,6]=="T")
signature_counts_noCtoT <- signature_counts[,-indices];
signature_set_noCtoT <- signature_set[-indices];
```

## classtpx with modern cluster

### K=2

```{r}

class_labs <- rep(1,25)
known_samples <- 26:50

Topic_clus <- classtpx::class_topics(
  signature_counts_noCtoT, 
  K=2, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.5,0.5),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

### K=3

Now for K=3

```{r}
Topic_clus <- class_topics(
  signature_counts_noCtoT, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.9999,0.001,0.001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set_noCtoT[x])

imp_features

```

## classtpx with uniform cluster

### K=3

Now we take one of the clusters to have equal weight on all damage signatures.
We first take K=3.

```{r}
Topic_clus <- class_topics(
  signature_counts_noCtoT, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/length(signature_set_noCtoT), length(signature_set_noCtoT)),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.9999,0.001,0.001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

### K=4

We assume one of the clusters to have equal weight on all damage signatures
Next we take K=4.

```{r}
Topic_clus <- class_topics(
  signature_counts_noCtoT, 
  K=4, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/length(signature_set_noCtoT), length(signature_set_noCtoT)),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.99,0.01,0.01,0.01),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

indices <- CountClust::ExtractTopFeatures(Topic_clus$theta, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set_noCtoT[x])

imp_features

```

