---
title: "Factor Analysis on Ancient Damage Data"
author: "Kushal K Dey"
date: "August 14, 2016"
output: html_document
---

```{r echo=FALSE, eval=TRUE}
Rcpp::sourceCpp("../../flashr/src/cloglik.cpp")
Rcpp::sourceCpp(("../../flashr/src/ATM.cpp"))
source("../../flashr/R/flash.R")
source("../../flashr/R/greedy.R")
source("../../flashr/R/FactorGGBar.R")
source("../../flashr/R/FactorGGplot.R")
source("../../flashr/R/flash_factor_postprocess.R")
source("../R/visDMsignature.R")
library(CountClust)
library(ggplot2)
library(pmsignature)

gsub2 <- function(pattern, replacement, x, ...) {
  for(i in 1:length(pattern))
    x <- gsub(pattern[i], replacement[i], x, ...)
  x
}
```

Load the Lindo 2016 Ancient Damage data

```{r echo=TRUE, eval=TRUE}
signature_counts <- get(load("../summary_data/signature-counts-clubbed-Lindo2016.rda"))

signature_set <- colnames(signature_counts)

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

site_left_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,1])
site_left_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,2])
site_right_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,7])
site_right_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,8])

sub_pattern <- sapply(1:dim(sig_split)[1], function(x) paste(sig_split[x,3:6], collapse=""))

from = c("C->A", "C->G", "C->T", "T->A", "T->C", "T->G")
to = c(1,2,3,4,5,6)

substitute <- gsub2(from, to, sub_pattern)

mutation_features <- sapply(1:dim(sig_split)[1], 
                                   function(m) paste0(substitute[m], ",", site_left_2[m], ",", site_left_1[m], ",", site_right_1[m], ",", site_right_2[m]))

mut_features_mat <- data.frame(cbind(substitute, site_left_2, site_left_1, site_right_1, site_right_2))
rownames(mut_features_mat) <- mutation_features
colnames(mut_features_mat)=NULL
mut_features_mat <- as.matrix(apply(mut_features_mat, c(1,2), function(x) as.numeric(x)))

```

We briefly explore the data summary.

```{r echo=TRUE, eval=TRUE}
## dimensions of data: nrows=number of samples 
## ncols = number of damage/mutation signatures

print(dim(signature_counts))

## the top corner of the counts matrix

signature_counts[1:5,1:5]
```

We convert the counts data to log CPM data by the `limma::voom` transformation.

```{r echo=TRUE, eval=TRUE}
### voom transformation

voom_signature_counts <- t(limma::voom(t(signature_counts))$E);

### track the voom weights 

voom_weights <- t(limma::voom(t(signature_counts))$weights);

write.table(voom_signature_counts, file="../../sfa_seq/sfa_inputs/sfa_input_voom_sig_with_CtoT_lindo2016.txt", row.names=FALSE,col.names=FALSE, quote=FALSE, sep="\t")

```


### Principal Component Analysis

```{r echo=FALSE, eval=TRUE}
pr <- prcomp(voom_signature_counts)

pc_data_frame <- data.frame("PC"=pr$x,
                            "labels"=c(rep("Ancient",25),
                                       rep("Modern",25)))

graphList <- vector(mode="list");
library(ggplot2)


graphList[[1]] <- qplot(PC.PC1, PC.PC2,
      data=pc_data_frame,
      colour=labels)

graphList[[2]] <-qplot(PC.PC1, PC.PC3,
      data=pc_data_frame,
      colour=labels)

graphList[[3]] <-qplot(PC.PC2, PC.PC3,
      data=pc_data_frame,
      colour=labels)

library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 2,
                      nrow = 2))


```

## Exploring substitution patterns 

```{r}
sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

sig_split_red <- sig_split[,3:6];

substitute_sig <- sapply(1:dim(sig_split_red)[1], function(x) paste0(sig_split_red[x,], collapse=""))

normalize <- function(x) return(x/sum(x))
prop_substitute_sig <- do.call(rbind, 
          lapply(1:dim(signature_counts)[1], function(x) 
          normalize(tapply(signature_counts[x,], substitute_sig, sum))))

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(prop_substitute_sig))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(prop_substitute_sig) <- annotation$sample_id;

StructureGGplot(omega = prop_substitute_sig,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Development Phase",
                order_sample = FALSE,
                figure_title = paste0("StructurePlot: 6 substituion patterns (with C -> T)"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

```

We now load the FLASH results for the data.

```{r echo=TRUE, eval=FALSE}
ll <- flashpool(voom_signature_counts, K=10, 
                 tol=1e-3, maxiter_r1 = 100,
                 partype="constant", 
                 nonnegative=FALSE)

save(ll, file="../rda/greedy_flash_voom_signature_Lindo2016.rda")
```

```{r echo=TRUE, eval=TRUE}
ll <- get(load("../rda/greedy_flash_voom_signature_Lindo2016.rda"))

postprocess_ll <- flash_factor_postprocess(ll$l,ll$f, voom_signature_counts)
pve_percentage <- c("", paste0(": PVE : ", round(postprocess_ll$PVE*100,1), "%"))

```

```{r}
library(ggplot2)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(ll$l))),
  label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(ll$l) <- paste0("X", c(1:NROW(ll$l)));

FactorGGBar(loadings = ll$l,
            annotation = annotation,
            palette = list("mid"="white", 
                           "low"="red", 
                           "high"="blue", 
                           "midpoint"=0),
             yaxis_label = "Population Type",
             figure_title = " ",
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 7,
                              axis_label_face = "bold"),
            legend_labels = pve_percentage,
            scale=TRUE,
            panel=list(panel_rows=2,
                       panel_title="greedy FLASH Factor Loadings Bar plot (with C to T)",
                       panel_title_fontsize=10,
                       panel_title_font=3))

```

```{r}
FactorGGplot(loadings = ll$l[,-1],
             annotation = annotation,
             palette = c( RColorBrewer::brewer.pal(8, "Accent"),
                          RColorBrewer::brewer.pal(4, "Spectral")),
             figure_title = "greedy FLASH Factor loadings stacked barchart (with C to T)",
             yaxis_label = "Factor type",
             order_sample = FALSE,
             sample_order_decreasing = TRUE,
             split_line = list(split_lwd = 0.2,
                               split_col = "black"),
             plot_labels = TRUE,
             legend_labels = pve_percentage[-1],
             scale=TRUE,
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 3,
                              axis_label_face = "bold"))

```

```{r}
rownames(ll$f) <- signature_set
ll_f_scale <- apply(ll$f[,-1],2,function(x) 
                                  {
                                      if(sd(x)!=0) {return (x/sd(x))}
                                      else {return (x)}
    })

normalize <- function(x) { return (x/sum(x))}

abs_f_scale <- apply(ll_f_scale,2,function(x) normalize(abs(x)))
indices <- CountClust::ExtractTopFeatures(abs_f_scale, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

## Sparse Factor Analysis

We now apply SFA and compare the results.

```{r}
# ./sfa_mac -gen ../../sfa_inputs/voom_sig_with_CtoT_lindo2016.txt -g 50 -n 1536 -k 10 -iter 500 -r 800 -o ../../sfa_outputs/Lindo2016/without_CtoT/voom_lindo_without_CtoT
#  ./sfa_mac -gen ../../sfa_inputs/sfa_input_voom_sig_with_CtoT_lindo2016.txt -g 50 -n 1536 -k 10 -iter 500 -r 800  -o ../../sfa_outputs/Lindo2016/with_CtoT/voom_lindo_with_CtoT

```


```{r}
loadings <- read.table("../utilities/sfa_Lindo2016/with_CtoT/voom_lindo_with_CtoT_lambda.out")
factors <- read.table("../utilities/sfa_Lindo2016/with_CtoT/voom_lindo_with_CtoT_F.out")
ll <- list("l"=loadings, "f"=t(factors))

postprocess_ll <- flash_factor_postprocess(ll$l,ll$f, voom_signature_counts)
pve_percentage <- c("", paste0(": PVE : ", round(postprocess_ll$PVE*100,1), "%"))
```

```{r}
library(ggplot2)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(ll$l))),
  label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(ll$l) <- paste0("X", c(1:NROW(ll$l)));

FactorGGBar(loadings = ll$l,
             annotation = annotation,
            palette = list("mid"="white", 
                           "low"="red", 
                           "high"="blue", 
                           "midpoint"=0),
             yaxis_label = "Population Type",
             figure_title = " ",
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 7,
                              axis_label_face = "bold"),
            legend_labels = pve_percentage,
            scale=TRUE,
            panel=list(panel_rows=2,
                       panel_title="SFA Factor Loadings Bar plot (with C to T)",
                       panel_title_fontsize=10,
                       panel_title_font=3))

```

```{r}
FactorGGplot(loadings = ll$l[,-1],
             annotation = annotation,
             palette = c( RColorBrewer::brewer.pal(8, "Accent"),
                          RColorBrewer::brewer.pal(4, "Spectral")),
             figure_title = "SFA Factor loadings stacked barchart (with C to T)",
             yaxis_label = "Factor type",
             order_sample = FALSE,
             sample_order_decreasing = TRUE,
             split_line = list(split_lwd = 0.2,
                               split_col = "black"),
             plot_labels = TRUE,
             legend_labels = pve_percentage[-1],
             scale=TRUE,
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 3,
                              axis_label_face = "bold"))

```

```{r}
rownames(ll$f) <- signature_set
ll_f_scale <- apply(ll$f[,-1],2,function(x) 
                                  {
                                      if(sd(x)!=0) {return (x/sd(x))}
                                      else {return (x)}
    })

normalize <- function(x) { return (x/sum(x))}

abs_f_scale <- apply(ll_f_scale,2,function(x) normalize(abs(x)))
indices <- CountClust::ExtractTopFeatures(abs_f_scale, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```

## Penalized Matrix Decomposition

```{r echo=TRUE, eval=FALSE}
ll_pmd <- PMA::PMD(voom_signature_counts, 
               type="standard",
               K=10)
ll <- list("l"=ll_pmd$u, "f"=ll_pmd$v)
postprocess_ll <- flash_factor_postprocess(ll$l,ll$f, voom_signature_counts)
pve_percentage <- c("", paste0(": PVE : ", round(postprocess_ll$PVE*100,1), "%"))


```

```{r}
library(ggplot2)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(ll$l))),
  label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(ll$l) <- paste0("X", c(1:NROW(ll$l)));

FactorGGBar(loadings = ll$l,
             annotation = annotation,
            palette = list("mid"="white", 
                           "low"="red", 
                           "high"="blue", 
                           "midpoint"=0),
             yaxis_label = "Population Type",
             figure_title = " ",
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 7,
                              axis_label_face = "bold"),
            scale=TRUE,
            panel=list(panel_rows=2,
                       panel_title="PMD Factor Loadings Bar plot (with C to T)",
                       panel_title_fontsize=10,
                       panel_title_font=3))

```

```{r}
FactorGGplot(loadings = ll$l,
             annotation = annotation,
             palette = c( RColorBrewer::brewer.pal(8, "Accent"),
                          RColorBrewer::brewer.pal(4, "Spectral")),
             figure_title = "PMD Factor loadings stacked barchart (with C to T)",
             yaxis_label = "Factor type",
             order_sample = FALSE,
             sample_order_decreasing = TRUE,
             split_line = list(split_lwd = 0.2,
                               split_col = "black"),
             plot_labels = TRUE,
             legend_labels = "",
             scale=TRUE,
             axis_tick = list(axis_ticks_length = .1,
                              axis_ticks_lwd_y = .1,
                              axis_ticks_lwd_x = .1,
                              axis_label_size = 3,
                              axis_label_face = "bold"))

```

```{r}
rownames(ll$f) <- signature_set
ll_f_scale <- apply(ll$f,2,function(x) 
                                  {
                                      if(sd(x)!=0) {return (x/sd(x))}
                                      else {return (x)}
    })

normalize <- function(x) { return (x/sum(x))}

abs_f_scale <- apply(ll$f,2,function(x) normalize(abs(x)))
indices <- CountClust::ExtractTopFeatures(abs_f_scale, top_features=10, method="poisson", options="min")
imp_features <- apply(indices, c(1,2), function(x) signature_set[x])

imp_features

```


## CountCluster application

```{r}
topics_clus <- get(load("../rda/CountClust_output_Lindo2016_with_C_to_T.rda"))
```

### K=2

```{r}
theta <- topics_clus$clust_2$theta;
omega <- topics_clus$clust_2$omega

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
sample_names <- rownames(signature_counts)
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))


K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}


Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 2))

omega <- slot(Param, "sampleSignatureDistribution")
library(CountClust)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Development Phase",
                order_sample = FALSE,
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])


```


### K=3

```{r}
theta <- topics_clus$clust_3$theta;
omega <- topics_clus$clust_3$omega

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 3))

omega <- slot(Param, "sampleSignatureDistribution")
library(CountClust)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Development Phase",
                order_sample = FALSE,
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])


```

### K=4

```{r}
theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)


K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
graphList[[4]] <- visPMSignature(Param, 4)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 2,
                      nrow = 2))

omega <- slot(Param, "sampleSignatureDistribution")
library(CountClust)
annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Ancient",25), rep("Modern",25)))
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Development Phase",
                order_sample = FALSE,
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: without C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])


```

```{r}
theta <- topics_clus$clust_3$theta;
omega <- topics_clus$clust_3$omega

sample_names <- rownames(signature_counts)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))

visDMSignature (theta, pattern,
                sample_names,numBases, flankingBasesNum,
                trDir, fdim, title_size=10,   panel_title_size=10, panel_title_font=4, layout=c(6,1))

```

