---
title: "Separate analysis (IO, RISE and Lindo)"
author: "Kushal K Dey"
date: "9/30/2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Packages + Scripts

We load the packages and source the scripts we would need for the analysis of the 152 ancients data.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
## the above scripts will eventually be part of flashr package
source("../R/visDMsignature.R")

## will eventually be part of the damage signature package

##  The libraries we need 

library(CountClust)
library(ggplot2)
library(pmsignature)
library(flashr)
```


Some helper functions we need 

```{r echo=TRUE, eval=TRUE}
gsub2 <- function(pattern, replacement, x, ...) {
  for(i in 1:length(pattern))
    x <- gsub(pattern[i], replacement[i], x, ...)
  x
}
```

## IO data analysis 

### Data Loading + Preprocessing

```{r echo=TRUE, eval=TRUE}
signature_counts_1 <- get(load("../summary_data/153-ancient-signature-counts-clubbed.rda"))

RISE_labs <- grep("RISE", as.character(rownames(signature_counts_1)))

signature_counts <- signature_counts_1[-RISE_labs,];

signature_set <- colnames(signature_counts)

## dimensions of data: nrows=number of samples 
## ncols = number of damage/mutation signatures

print(dim(signature_counts))

## the top corner of the counts matrix

signature_counts[1:5,1:5]
```

### PCA

Apply PCA 

```{r}
voom_signature_counts <- t(limma::voom(t(signature_counts))$E);
voom_weights <- t(limma::voom(t(signature_counts))$weights);

pr <- prcomp(voom_signature_counts)

pc_data_frame <- data.frame("PC"=pr$x,
                          "labels"=c(rep("IO",dim(signature_counts)[1])))

graphList <- vector(mode="list");
library(ggplot2)


graphList[[1]] <- qplot(PC.PC1, PC.PC2,
      data=pc_data_frame,
      colour=labels)

graphList[[2]] <-qplot(PC.PC1, PC.PC3,
      data=pc_data_frame,
      colour=labels)

graphList[[3]] <-qplot(PC.PC2, PC.PC3,
      data=pc_data_frame,
      colour=labels)

library(grid)
library(gridExtra)
a <- do.call("grid.arrange", 
               args = list(grobs=graphList,
                            ncol = 2,
                            nrow = 2))

```

### GoM model + Structure

```{r echo=TRUE, eval=FALSE}
topics_clus <- CountClust::FitGoM(signature_counts, K=2:4, tol=0.01, control=list(tmax=10000));
save(topics_clus, file="../rda/topics_clus_IO_2_to_4.rda")
```

```{r}
topics_clus <- get(load("../rda/topics_clus_IO_2_to_4.rda"))
```

```{r}
omega <- topics_clus$clust_2$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_2$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```


```{r}
omega <- topics_clus$clust_3$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_3$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```

```{r}
omega <- topics_clus$clust_4$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_4$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```

### Damage signatures

```{r}

## create a list of damage or mutation signatures

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

## convert the bases to numbers

site_left_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,1])
site_left_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,2])
site_right_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,7])
site_right_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,8])

sub_pattern <- sapply(1:dim(sig_split)[1], function(x) paste(sig_split[x,3:6], collapse=""))

from = c("C->A", "C->G", "C->T", "T->A", "T->C", "T->G")
to = c(1,2,3,4,5,6)

## pattern of bases matrix

substitute <- gsub2(from, to, sub_pattern)

mutation_features <- sapply(1:dim(sig_split)[1], 
                                   function(m) paste0(substitute[m], ",", site_left_2[m], ",", site_left_1[m], ",", site_right_1[m], ",", site_right_2[m]))

## matrix of mutation or damage features

mut_features_mat <- data.frame(cbind(substitute, site_left_2, site_left_1, site_right_1, site_right_2))
rownames(mut_features_mat) <- mutation_features
colnames(mut_features_mat)=NULL
mut_features_mat <- as.matrix(apply(mut_features_mat, c(1,2), function(x) as.numeric(x)))

```

```{r}
theta <- topics_clus$clust_2$theta;
omega <- topics_clus$clust_2$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 2))

```

```{r}
theta <- topics_clus$clust_3$theta;
omega <- topics_clus$clust_3$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 3))

```

```{r}
theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
graphList[[4]] <- visPMSignature(Param, 4)

library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 4))

```

### Conditional damage signatures

conditional damage 

```{r message=FALSE, warning=FALSE}
normalize <- function(x) return(x/sum(x))

theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

sample_names <- rownames(signature_counts)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))

index_sub <- c("C->T", "C->A", "C->G", "T->A", "T->C", "T->G")

visDMSignature (theta, 
                index_sub,
                pattern=sub_pattern,
                mutation_features_matrix = mut_features_mat,
                sample_names,
                flankingBasesNum=as.integer(numBases),
                trDir, 
                fdim, 
                title_size=10,   
                panel_title_size=10, 
                panel_title_font=4, 
                layout=c(6,1))

```


## RISE data

### Data Loading + Preprocessing

```{r}
signature_counts <- signature_counts_1[RISE_labs,];

signature_set <- colnames(signature_counts)

## dimensions of data: nrows=number of samples 
## ncols = number of damage/mutation signatures

print(dim(signature_counts))

## the top corner of the counts matrix

signature_counts[1:5,1:5]

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

indices <- which(sig_split[,3]=="C" & sig_split[,6]=="T")

signature_counts_noCtoT <- signature_counts[,-indices];

signature_set_noCtoT <- signature_set[-indices];


voom_signature_counts_noCtoT <- t(limma::voom(t(signature_counts_noCtoT))$E);


### track the voom weights 

voom_weights_noCtoT <- t(limma::voom(t(signature_counts_noCtoT))$weights);


```

### PCA

Apply PCA 

```{r}

pr <- prcomp(voom_signature_counts_noCtoT)

pc_data_frame <- data.frame("PC"=pr$x,
                     "labels"=c(rep("RISE",dim(signature_counts_noCtoT)[1])))

graphList <- vector(mode="list");
library(ggplot2)


graphList[[1]] <- qplot(PC.PC1, PC.PC2,
      data=pc_data_frame,
      colour=labels)

graphList[[2]] <-qplot(PC.PC1, PC.PC3,
      data=pc_data_frame,
      colour=labels)

graphList[[3]] <-qplot(PC.PC2, PC.PC3,
      data=pc_data_frame,
      colour=labels)

library(grid)
library(gridExtra)
a <- do.call("grid.arrange", 
               args = list(grobs=graphList,
                            ncol = 2,
                            nrow = 2))

```

### GoM model + Structure

```{r echo=TRUE, eval=FALSE}
topics_clus <- CountClust::FitGoM(signature_counts_noCtoT, K=2:4, tol=0.01, control=list(tmax=10000));
save(topics_clus, file="../rda/topics_clus_RISE_2_to_4_noCtoT.rda")
```

```{r}
topics_clus <- get(load("../rda/topics_clus_RISE_2_to_4_noCtoT.rda"))
```

```{r}
omega <- topics_clus$clust_2$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_2$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set_noCtoT[x])

```


```{r}
omega <- topics_clus$clust_3$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_3$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set_noCtoT[x])

```

```{r}
omega <- topics_clus$clust_4$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_4$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```

### Damage Signatures

```{r}

## create a list of damage or mutation signatures

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

## convert the bases to numbers

site_left_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,1])
site_left_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,2])
site_right_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,7])
site_right_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,8])

sub_pattern <- sapply(1:dim(sig_split)[1], function(x) paste(sig_split[x,3:6], collapse=""))

from = c("C->A", "C->G", "C->T", "T->A", "T->C", "T->G")
to = c(1,2,3,4,5,6)

## pattern of bases matrix

substitute <- gsub2(from, to, sub_pattern)

mutation_features <- sapply(1:dim(sig_split)[1], 
                                   function(m) paste0(substitute[m], ",", site_left_2[m], ",", site_left_1[m], ",", site_right_1[m], ",", site_right_2[m]))

## matrix of mutation or damage features

mut_features_mat <- data.frame(cbind(substitute, site_left_2, site_left_1, site_right_1, site_right_2))
rownames(mut_features_mat) <- mutation_features
colnames(mut_features_mat)=NULL
mut_features_mat <- as.matrix(apply(mut_features_mat, c(1,2), function(x) as.numeric(x)))

sub_pattern_noCtoT <- sub_pattern[-indices];
mut_features_mat_noCtoT <- mut_features_mat[-indices,];


```

```{r}
theta <- topics_clus$clust_2$theta;
omega <- topics_clus$clust_2$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat_noCtoT[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 2))

```

```{r}
theta <- topics_clus$clust_3$theta;
omega <- topics_clus$clust_3$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat_noCtoT[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 3))

```

```{r}
theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat_noCtoT[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
graphList[[4]] <- visPMSignature(Param, 4)

library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 4))

```

### Conditional damage signatures

conditional damage 

```{r message=FALSE, warning=FALSE}
normalize <- function(x) return(x/sum(x))

theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

sample_names <- rownames(signature_counts_noCtoT)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))

index_sub <- c("C->A", "C->G", "T->A", "T->C", "T->G")

visDMSignature (theta, 
                index_sub,
                pattern=sub_pattern_noCtoT,
                mutation_features_matrix = mut_features_mat_noCtoT,
                sample_names,
                flankingBasesNum=as.integer(numBases),
                trDir, 
                fdim, 
                title_size=10,   
                panel_title_size=10, 
                panel_title_font=4, 
                layout=c(5,1))

```

## Lindo 2016

### Data Loading + Preprocessing

```{r echo=TRUE, eval=TRUE}
signature_counts <- get(load("../summary_data/signature-counts-clubbed-Lindo2016.rda"))

signature_counts <- signature_counts[1:25,]

signature_set <- colnames(signature_counts)

## dimensions of data: nrows=number of samples 
## ncols = number of damage/mutation signatures

print(dim(signature_counts))

## the top corner of the counts matrix

signature_counts[1:5,1:5]
```

### GoM model + Structure

```{r echo=TRUE, eval=FALSE}
topics_clus <- CountClust::FitGoM(signature_counts, K=2:4, tol=0.01, control=list(tmax=10000));
save(topics_clus, file="../rda/topics_clus_Lindo_ancient_2_to_4.rda")
```

```{r}
topics_clus <- get(load("../rda/topics_clus_Lindo_ancient_2_to_4.rda"))
```

```{r}
omega <- topics_clus$clust_2$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_2$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```


```{r}
omega <- topics_clus$clust_3$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_3$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```

```{r}
omega <- topics_clus$clust_4$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(1:dim(omega)[1])
)

rownames(omega) <- annotation$sample_id;

StructureGGplot(omega = omega,
                annotation = annotation,
                palette = RColorBrewer::brewer.pal(8, "Accent"),
                yaxis_label = "Moderns vs Ancients",
                order_sample = FALSE,
                split_line = list(split_lwd = 0.5,
                                  split_col = "white"),
                figure_title = paste0("StructurePlot: K=", dim(omega)[2],": pmsignature: with C->T/G->A"),
                axis_tick = list(axis_ticks_length = .1,
                                 axis_ticks_lwd_y = .1,
                                 axis_ticks_lwd_x = .1,
                                 axis_label_size = 7,
                                 axis_label_face = "bold"))

apply(ExtractTopFeatures(topics_clus$clust_4$theta, top_features = 10, method="poisson", options="min"), c(1,2), function(x) signature_set[x])

```

### Damage signature

```{r}

## create a list of damage or mutation signatures

sig_split <- do.call(rbind, lapply(colnames(signature_counts), function(x) strsplit(as.character(x), split="")[[1]]))

## convert the bases to numbers

site_left_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,1])
site_left_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,2])
site_right_1 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,7])
site_right_2 <- gsub2(c("A","C","G","T"), c(1,2,3,4), x=sig_split[,8])

sub_pattern <- sapply(1:dim(sig_split)[1], function(x) paste(sig_split[x,3:6], collapse=""))

from = c("C->A", "C->G", "C->T", "T->A", "T->C", "T->G")
to = c(1,2,3,4,5,6)

## pattern of bases matrix

substitute <- gsub2(from, to, sub_pattern)

mutation_features <- sapply(1:dim(sig_split)[1], 
                                   function(m) paste0(substitute[m], ",", site_left_2[m], ",", site_left_1[m], ",", site_right_1[m], ",", site_right_2[m]))

## matrix of mutation or damage features

mut_features_mat <- data.frame(cbind(substitute, site_left_2, site_left_1, site_right_1, site_right_2))
rownames(mut_features_mat) <- mutation_features
colnames(mut_features_mat)=NULL
mut_features_mat <- as.matrix(apply(mut_features_mat, c(1,2), function(x) as.numeric(x)))

```

```{r}
theta <- topics_clus$clust_2$theta;
omega <- topics_clus$clust_2$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 2))

```

```{r}
theta <- topics_clus$clust_3$theta;
omega <- topics_clus$clust_3$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 3))

```

```{r}
theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

library(pmsignature)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))
sample_names <- rownames(signature_counts)

K <- dim(omega)[2]
F <- array(0, c(K, length(fdim), max(fdim)))
for(k in 1:K){
  for(kk in 1:length(fdim)){
    temp <- tapply(theta[,k], mut_features_mat[,kk], sum)
    F[k,kk,as.numeric(names(temp))] <- as.numeric(temp)
  }
}

Param <- new(Class = "EstimatedParameters", 
           type = type,
           flankingBasesNum = flankingBasesNum,
           transcriptionDirection = trDir,
           possibleFeatures = as.integer(fdim),
           sampleList = sample_names,
           signatureNum = as.integer(K),
           isBackGround = isBG,
           signatureFeatureDistribution = F,
           sampleSignatureDistribution = omega,
           loglikelihood = -100000)

graphList <- list(mode="vector")
graphList[[1]] <- visPMSignature(Param, 1)
graphList[[2]] <- visPMSignature(Param, 2)
graphList[[3]] <- visPMSignature(Param, 3)
graphList[[4]] <- visPMSignature(Param, 4)

library(grid)
library(gridExtra)
do.call("grid.arrange", 
          args = list(grobs=graphList,
                      ncol = 1,
                      nrow = 4))

```

### Conditional damage signatures

conditional damage 

```{r message=FALSE, warning=FALSE}
normalize <- function(x) return(x/sum(x))

theta <- topics_clus$clust_4$theta;
omega <- topics_clus$clust_4$omega

sample_names <- rownames(signature_counts)

isBG <- TRUE
numBases <- 5
type <- "independent"
flankingBasesNum = as.integer(numBases)
trDir <- FALSE
fdim <- c(6, rep(4, numBases - 1), rep(2, as.integer(trDir)))

index_sub <- c("C->T", "C->A", "C->G", "T->A", "T->C", "T->G")

visDMSignature (theta, 
                index_sub,
                pattern=sub_pattern,
                mutation_features_matrix = mut_features_mat,
                sample_names,
                flankingBasesNum=as.integer(numBases),
                trDir, 
                fdim, 
                title_size=10,   
                panel_title_size=10, 
                panel_title_font=4, 
                layout=c(6,1))

```

