---
title: "Testing topic models on rare damage patterns"
author: "Kushal K Dey"
date: "September 14, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Introduction

We test whether GoM model can capture accurately  differences in grades of membership when one of the clusters represents common patterns and another represents rare patterns.

Load the packages 

```{r}
library(maptpx)
library(classtpx)
library(slam)
library(CountClust)
```


## K=3 Simulation

We present a simulation pattern for $K=3$. There are 6 features corresponding to 6 types of mutational patterns for the substituting base. 

### Design

```{r}
omega_modern <- c(1.0, 0, 0);
omega_ancient_1 <- c(0.7, 0.2, 0.1);
omega_ancient_2 <- c(0.7, 0.1, 0.2);

theta1 <- rep(1/6,6);  ## common pattern - equal weights on all damages 
theta2 <- c(1,rep(0,5)) ## rare pattern- first mutation pattern gets the weight 
theta3 <- c(rep(0,5),1) ## rare pattern - last mutation pattern gets the weight
```


```{r echo=FALSE, eval=TRUE}
rep.row<-function(x,n){
  matrix(rep(x,each=n),nrow=n)
}
rep.col<-function(x,n){
  matrix(rep(x,each=n), ncol=n, byrow=TRUE)
}
```


### Counts simulation

```{r}
omega_modern_set <- rep.row(omega_modern,25);
omega_ancient_1_set <- rep.row(omega_ancient_1, 20);
omega_ancient_2_set <- rep.row(omega_ancient_2, 5);

omega_set <- rbind(omega_modern_set,
                   omega_ancient_1_set,
                   omega_ancient_2_set)

theta_set <- cbind(theta1, theta2, theta3)

p_set <- omega_set%*%t(theta_set)

counts <- matrix(0, dim(p_set)[1], dim(p_set)[2])

for(n in 1:dim(counts)[1]){
  counts[n,] <- rmultinom(1, 100, prob=p_set[n,])
}
```

### maptpx model

```{r}
topics_clus <- maptpx::topics(counts, K=3, tol=0.001)

w <- topics_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(w))),
  tissue_label = factor(c(rep("Modern",25), rep("Ancient",25)))
)

rownames(w) <- annotation$sample_id;

CountClust::StructureGGplot(omega = w,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))


```

### classtpx (model 1)

classtpx model fit with one cluster fixed to be all equal damage signatures and 2 other clusters unknown.

```{r}
known_samples <- 1:25;
class_labs <- rep(1,25);

Topic_clus <- class_topics(
  counts, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/6,6),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.9999,0.0001,0.0001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Modern",25), rep("Ancient",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

### classtpx (model 2)

Now a classtpx model with one class fixed by the modern samples, and 2 unknown classes. 

```{r}
known_samples <- 1:25
class_labs <- rep(1,25)

Topic_clus <- class_topics(
  counts, 
  K=3, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.9999,0.0001,0.0001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Modern",25), rep("Ancient",25)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```


## K=2 simulation

### Design

```{r}
omega_modern <- c(0.9,0.1)
omega_ancient_1 <- c(0.75,0.25);

theta1 <- rep(1/6,6); ## equal prob to 6 types of mutations
theta2 <- c(1,rep(0,5))
```

### Counts Simulation 

```{r}
omega_modern_set <- rep.row(omega_modern,200);
omega_ancient_1_set <- rep.row(omega_ancient_1, 200);

omega_set <- rbind(omega_modern_set,
                   omega_ancient_1_set)

theta_set <- cbind(theta1, theta2)

p_set <- omega_set%*%t(theta_set)

counts <- matrix(0, dim(p_set)[1], dim(p_set)[2])

for(n in 1:dim(counts)[1]){
  counts[n,] <- rmultinom(1, 100, prob=p_set[n,])
}

```

### maptpx 

```{r}
topics_clus <- maptpx::topics(counts, K=2, tol=0.01)

w <- topics_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(w))),
  tissue_label = factor(c(rep("Modern",200), rep("Ancient",200)))
)

rownames(w) <- annotation$sample_id;

CountClust::StructureGGplot(omega = w,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

### classtpx (model 1)

```{r}
known_samples <- 1:200;
class_labs <- rep(1,200);
library(maptpx)
library(slam)

Topic_clus <- class_topics(
  counts, 
  K=2, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(1,0.001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Modern",200), rep("Ancient",200)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

### classtpx (model 2)

```{r}
known_samples <- 1:200;
class_labs <- rep(1,200);

Topic_clus <- class_topics(
  counts, 
  K=2, 
  known_samples = known_samples,
  class_labs = class_labs,
  method="theta.fix",
  optional_theta = rep(1/6,6),
  shrink=FALSE,
  tol=0.01,
  prior_omega = c(0.999999,0.000001),
  ord=FALSE)

omega <- Topic_clus$omega

annotation <- data.frame(
  sample_id = paste0("X", c(1:NROW(omega))),
  tissue_label = factor(c(rep("Modern",200), rep("Ancient",200)))
)

rownames(omega) <- annotation$sample_id;

CountClust::StructureGGplot(omega = omega,
                            annotation = annotation,
                            palette = RColorBrewer::brewer.pal(8, "Accent"),
                            yaxis_label = "Moderns vs Ancients",
                            order_sample = FALSE,
                            figure_title = "StructurePlot",
                            axis_tick = list(axis_ticks_length = .1,
                                             axis_ticks_lwd_y = .1,
                                             axis_ticks_lwd_x = .1,
                                             axis_label_size = 7,
                                             axis_label_face = "bold"))

```

